---
title: 'Unit 7 : Statistical Testing'
author: "Te Rutherford"
date: "November 19, 2014"
output: ioslides_presentation
---

## Long lunch vs short lunch data 

There are a few employees in the short lunch break group who are more productive than the ones in the long lunch break group. Why? 

```{r, echo=FALSE}
data = read.csv('http://www.typ-stats.com/datasets/lunch_data.csv')
head(subset(data, condition =='long'))
head(subset(data, condition =='short'))
```

## Statistical noise

There are at least three sources of noise:

* Data collection error
* Inherent variation in the system 
* Error caused by sampling

The noise in the data make us uncertain about the conclusion we can draw. These are the very reasons why we need to use statistical tests. 

## Statistical tests

Statistical tests help indicate whether we have enough signal (and little enough noise) to make a conclusion.
There are four steps in conducting statistical test

* Formulate the question
* Re-formulate the question mathematically
* Compute the probability distribution over the test statistic
* Draw a conclusion from the probability distribution


Let’s look at the overview of each step

## 1. Formulate the question

From the electronic gadget dataset, for example, we would like to answer this multiple-choice question:
Do longer lunch breaks increase the number of gadgets that employees can produce?

* Yes. Long lunch employees produce more.
* No. Long lunch employees produce less.

## 2. Re-formulate the question mathematically

Let $\bar{X_1}$ and $\bar{X_2}$ be the *true* mean number of gadgets produced by long lunch employees and short lunch employees respectively. Then let $Y=\bar{X_1} - \bar{X_2}$ be the difference between the two means. 

$Y$ is called a test statistic, the measure that will decide the outcome of the statistical test.
Which of the following is true?

* $Y>0$ Long lunch employees produce significantly more.
* $Y<0$ Long lunch employees produce significantly less.

Remember our motivation for statistical hypothesis testing? 


## 3. Compute the probability distribution over the test statistic

Suppose we have ways to compute $P(Y)$

For this running example, we need to compute 

* The probability that $Y>0$ (long lunch employees significantly produce more) = 0.98
* The probability that $Y<0$ (long lunch employees significantly produce less) = 0.02

This is the more sophisticated step in statistical tests. We will discuss this step in more detail.

## 4. Draw a conclusion from the probability distribution

We have to set a significance level, which is the level of confidence that we require until we draw a conclusion. It is conventionally set to 0.05 level. 

If not, then we cannot draw any (decisive) conclusion.

For our example, 

* The probability that $Y>0$ (long lunch employees produce significantly more) = 0.98
* The probability that $Y<0$ (long lunch employees produce significantly less) = 0.02

We conclude a. because the probability is greater than 0.95.  This is the answer to the original question posed in step 1. 

## 

Hypothetically, though if the results from step 3 give us. 

* The probability that $Y>0$ (long lunch employees produce significantly more) = 0.70
* The probability that $Y<0$ (long lunch employees produce significantly less) = 0.30


We say that we do not have enough evidence to support the hypothesis that long lunch breaks affect the productivity one way or another. 


## Significance level and the world

Let’s look at this case again.

* The probability that $Y>0$ (long lunch employees produce significantly more) = 0.70
* The probability that $Y<0$ (long lunch employees produce significantly less) = 0.30

We are 70% sure about a. 

* If we want to conclude a., do we have to raise or lower significance level?
* Why can’t we just adjust our significance level to be more level and conclude a.? 
* Suppose we do that, what will be the consequences?

# Bootstrap test


##  Bootstrap for constructing a probability distribution over a test statistic

Bootstrap is a simple technique for computing the probability of the test statistic taking a certain range of value. To demonstrate how each step works, let’s go through an example from the soporific drug study. 

* Before we start, go over the four steps for conducting hypothesis testing for this example.

\pagebreak

# Bootstrap algorithm (steps of calculation)

## Step 1. Resample the data

By resampling, one take random rows from the dataset to make a fake dataset of the same size. We call this bootstrap resample. For example, if the original dataset looks like this

```{r}
sleep
```

## 
Then a fake dataset (bootstrap resample) might look like this: 
```{r}
new_data = sleep[sample(1:nrow(sleep), replace=TRUE),]
new_data
```


## Step 2. Compute the quantity of interest from the bootstrap resample.

We are interested in the mean difference in extra hours of sleep. So we will go ahead and compute that the mean difference. 

```{r} 
group1_data = subset(new_data, group == 1) 
group2_data = subset(new_data, group == 2)
mean(group1_data$extra) - mean(group2_data$extra)
```

## Step 3. Repeat steps 1 and 2 many many times. 

In this dataset, we will end up with many different values of the mean difference in extra hours of sleep (computed from the bootstrap resamples).


From these values, we can compute the probability that the mean difference in extra hours of sleep is greater than zero.  = 


$$\frac{ \mbox{the number of bootstrap resamples whose mean difference} > 0} 
      {\mbox{the total number of bootstrap resamples}}$$

## Bootstrap in R 

All of this can be done using package boot in R.

```{r}
library(bootstrap)    
group1_means = bootstrap(group1_data$extra, 100, mean)$thetastar
group2_means = bootstrap(group2_data$extra, 100, mean)$thetastar
y = group1_means - group2_means
sum(y > 0)/ length(y)
```


\