---
title: "Confidence Interval of Proportion"
author: "Te Rutherford"
date: "January 29, 2015"
output: beamer_presentation
---

## Today

* We will look at confidence interval and what it means.

# What is confidence interval? and why do we care?

* Confidence interval gives you the uncertainty of our estimate. 
* When you say that a proportion is around 0.3 - 0.5, we want to make this more rigorous and formal. 
* We want to say the we are 90% confident that the true proportion is in the interval.

# Intuition

* We want to say that we are 90% confident that the true proportion is in 0.4 - 0.6.
* If you make the interval bigger, then you can be more confident.

# What do you mean by confidence?

* A 95% confidence interval means that if you run this survey/study/experiment many times 
and compute 95% confidence interval, then 95% of the intervals will cover the true proportion.
* Crap, this is so confusing. Let's look at an example.

# Example

* Suppose we are curious about the proportion of Brandeis students 
that had personal tutoring during junior in high school.

* We have to conduct a survey to estimate the proportion.
* Each survey conductor computes the estimated proportion and the 95% confidence interval.
* Look at the picture I drew on the board.

# How do we compute 95% confidence interval?

* Obviously, we use R.
* Suppose we are looking at the new medication data from last time. 
  +  If a new medication works on 300 patients but does not work on 150 patients,
the rate of medication being effective is 300 / (300 + 150)
* BTW, rate is now called proportion because Manny asked for it. But it is really interchangeable.

# Confidence interval in R

* We use `bcanon` function to compute confidence interval
* If we want (100 - a)% confidence interval, 
then the lower bound is at the a/2 percentile, and
the upper bound is at the 100 - (a/2) percentile.
```{r}
library(bootstrap)
medication_data = c(rep(0,150), rep(1,300))
b= bcanon(medication_data, nboot= 10000, theta=mean)
b$confpoints
```
* Then you need to look up in the table.

# Problem 7.54 in the handout
* The proportion of people satisfy MSDS = 11/100
* 150 people have been surveyed.
* First step: we want to create a data vector using `rep` command.
There are 150 * 11/100 successes and 150 * 89/100 failures.
```{r}
msds_data = c(rep(0, 134), rep(1, 16))
```

# Problem 7.54 in the handout (continued)
```{r,echo=FALSE}
set.seed(10)
```
```{r}
b= bcanon(msds_data, nboot= 20000, theta=mean)
b$confpoints
```
* For the confidence level we set (100 - 5%), the lower bound is $\frac{5}{2}$th percentile and the upper bound is $100-\frac{5}{2}$ percentile. 
* So the 95% confidence interval is [0.06, 0.16]. We are 95% confident that the true proportion is in this interval.

# Specify the percentile you want to begin with
```{r,echo=FALSE}
set.seed(10)
```
```{r}
b= bcanon(msds_data, nboot= 20000, theta=mean, 
          alpha = c(0.025, 0.975))
b$confpoints
```
* By putting in `alpha = c(0.025, 0.975)`, we only get the percentiles that we want.
* If you don't put this in, you might not get the percentiles you want.
