---
title: "Comparing proportions across group and Error bars"
author: "Te Rutherford"
date: "February 3, 2015"
output:
  beamer_presentation:
    fig_height: 3
    fig_width: 2
---

## Today

* Interpretation of confidence interval
* Inference with confidence interval
* Comparing rates/proportions from multiple groups
* Error bars

## Problem 7.54 in the handout
* The proportion of people satisfy MSDS = 11/100
* 150 people have been surveyed.
* First step: we want to create a data vector using `rep` command.
There are 150 * 11/100 successes and 150 * 89/100 failures.
```{r}
msds_data = c(rep(0, 134), rep(1, 16))
```

## Find 95% confidence interval of the proportion
```{r,echo=FALSE}
set.seed(10)
```
```{r}
library(bootstrap)
b= bcanon(msds_data, nboot= 20000, theta=mean, 
          alpha = c(0.025, 0.975))
b$confpoints
```
* By putting in `alpha = c(0.025, 0.975)`, we only get the percentiles that we want.
* If you don't put this in, you might not get the percentiles you want.

## Interpretation of Confidence Interval

* You should report the estimated proportion along with the confidence interval. 
    + The estimated proportion of people who satisfy MSDS is 0.10 (0.06, 0.16). 
    + The estimated proportion of people who satisfy MSDS is 0.10 ($\pm 0.06$).
* Theoretical interpretation
    + If we repeat this study many times and draw many 95% confidence intervals, 95% of the intervals will cover the true proportion.
* Practical interpretation
    + We are 95% confident that the true proportion is in the interval of (0.06, 0.16).

## Effects of sample size 

* In general, as we increase sample size, the confidence interval becomes smaller.
```{r}
library(bootstrap)
msds_data = c(rep(0, 134), rep(1, 16))
b= bcanon(msds_data, nboot= 20000, theta=mean, 
          alpha = c(0.025, 0.975))
b$confpoints
```

```{r}
library(bootstrap)
msds_data = c(rep(0, 1340), rep(1, 160))
b= bcanon(msds_data, nboot= 20000, theta=mean, 
          alpha = c(0.025, 0.975))
b$confpoints
```







## Inference with confidence interval

* We are 95% confident that the true proportion is not anything outside the interval of (0.06, 0.16)
* The true proportion is *significantly different* from 0.05. 

## What if we want to compare the proportions from two groups?







* New command for plotting the summary statistics (without `ddply`)

## Plot
```{r}
library(ggplot2)
data = read.csv(
  'http://www.typ-stats.com/datasets/RicciData.csv')
ggplot(data=data,mapping=aes(x=Race, y=Pass)) + 
  stat_summary(fun.y=mean, geom='bar')
```

## Error bar
Draw the bars first and then draw the confidence intervals (error bars).
```{r}
ggplot(data=data,mapping=aes(x=Race, y=Pass)) + 
  stat_summary(fun.y=mean, geom='bar',fill='pink')+
  stat_summary(fun.data=mean_cl_boot, geom='pointrange')
```

## Error bar (2)
This kind is a bit ugly.
```{r}
ggplot(data=data,mapping=aes(x=Race, y=Pass)) + 
  stat_summary(fun.y=mean, geom='bar') + 
  stat_summary(fun.data=mean_cl_boot, geom='errorbar')
```

