---
title: "Wrapping up Hypothesis Testing on Proportion"
author: "Te Rutherford"
date: "February 5, 2015"
output:
  beamer_presentation:
    fig_height: 3
    fig_width: 2
---

## Today

* One more example of confidence interval and hypothesis testing
* Practical significance vs statistical significance
* What if we have population data

## GT200 the bomb detector

* Probably the most expensive scam in the world with significant political repercussion.
* https://www.youtube.com/watch?v=Ykv4PU7QXnU

## Testing GT200

* Systematic test
* Four boxes one of which contains a bomb. Go find it using this plastic piece of junk.
* If you pick randomly, what is the rate of identifying the bomb?
* 60 trials of the test. The officers correctly identify the bomb 18 times. 

## Bootstrap hypothesis testing

```{r}
library(bootstrap)
gt200data = c(rep(1, 18), rep(0, 60-18))
b = bootstrap(gt200data, nboot=10000, theta=mean)
sum(b$thetastar > 0.25) / 10000
```
What do we conclude?

## Bootstrap confidence interval

```{r}
gt200data = c(rep(1, 18), rep(0, 60-18))
b = bcanon(gt200data, nboot=10000, theta=mean, 
           alpha=c(0.975, 0.025))
b$confpoints
```
What is the practical interpretation of the confidence interval?

