---
title: "Assessing the Utility of the Model"
author: "Te Rutherford"
date: "March 10, 2015"
output:
  beamer_presentation:
    fig_height: 2
    fig_width: 3
---
# Agenda
* Last time
    + Sum of Squared Error
    + Fitting the linear regression model with `lm`
* Today
    + Assessing the utility of the model

# Fitting the linear regression model 
* Method of Least Square is used by `lm` function. 
```{r}
d = read.csv('http://www.typ-stats.com/datasets/pressure.csv')
lm(compression ~ pressure, data=d)
```

# Should we need $\beta_1$?

$$ y = \beta_0 + \beta_1 x + \epsilon$$

* $\beta_0$ is called intercept term.
* $\beta_1$ is called the slope. 

$$ y = \beta_0  + \epsilon$$

* Just the intercept term 

# The P-value for the slope term

```{r}
d = read.csv('http://www.typ-stats.com/datasets/pressure.csv')
insanelinearmodel = lm(compression ~ pressure, data=d)
summary(insanelinearmodel)
```

# The P-value

* If p-value for the slope term is less than 0.05, then the slope term significantly contributes to the prediction of the dependent variable $y$ using the linear model. 

* The p-value is at 0.0354. It even has a star next to it.

# R-Squared or the coefficient of determination

* SSE is meaningful when comparing models.  (relative term)
* R-Squared or the coefficient of determination indicates how well the model does in the absolute term.

# What is R-Squared?

* The improvement on SSE from adding $\beta_1$. 
* Compute SSE from Model 1
$$ y = \beta_0 + \beta_1 x + \epsilon$$
* Compute SSE from Model 2
$$ y = \beta_0  + \epsilon$$

* Compute the percent reduction of SSE. 

# Understanding R-Squared

* What happens when R-Squared is low?
* What happens when R-Squared is high?

#10.58
a. 
library(ggplot2)
ggplot(data=d, mapping=aes(x=TEMP, y= PROPPASS)) + geom_point()

b. 
lm(PROPPASS ~ TEMP, data= d)
ggplot(data=d, mapping=aes(x=TEMP, y= PROPPASS)) + geom_point() + geom_smooth(method='lm')
ggplot(data=d, mapping=aes(x=TEMP, y= PROPPASS)) + geom_point() + geom_abline(intercept=-13.49, slope=-0.05)

crazylinearmodel = lm(PROPPASS ~ TEMP, data= d)
summary(crazylinearmodel)
