---
title: "03-17-15-multiple-regression"
author: "Te Rutherford"
date: "March 17, 2015"
output:
  beamer_presentation:
    fig_height: 2
    fig_width: 3
    incremental: yes
---
# Fitting multiple regression model

* The data look like this
```{r}
mydata = data.frame(pressure = 1:5, temperature=c(1,2,2,4,3), compression=c(1,1,2,2,4))
mydata
```

# R Command for fitting multiple regression
```{r,eval=FALSE}
nameyourmodelhere = lm(compression ~ pressure + temperature, 
                       data=mydata)
summary(nameyourmodelhere)
```
    all:
    lm(formula = compression ~ pressure + temperature, 
      data = mydata)

    Residuals:
        1     2     3     4     5 
     0.25 -0.15 -0.30  0.05  0.15 

    Coefficients:
                Estimate Std. Error t value Pr(>|t|)  
    (Intercept)   0.3500     0.3640   0.962   0.4377  
    pressure      1.1500     0.1803   6.379   0.0237 *
    temperature  -0.7500     0.2500  -3.000   0.0955 .
    ---
    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

    Residual standard error: 0.3162 on 2 degrees of freedom
    Multiple R-squared:  0.9667,  Adjusted R-squared:  0.9333 
    F-statistic:    29 on 2 and 2 DF,  p-value: 0.03333

# New concepts
* What is $r^2$?
* What is adjusted $r^2$? What is it adjusted for? 

# Steps in regression analysis

* Determine predictors ($x_1, x_2, \dots$) and dependent variable $y$
* Fit the linear regression model with `lm` command
* Interpret the regression coefficients ($\beta_0, \beta_1, \beta_2, \dots, \beta_n$)
* Interpret $r^2$ and adjusted $r^2$. (How good is our model?)
* Determine whether $\beta_1$ is significant. Interpret the significance. 


