---
title: "Unit 5 Part 1"
author: "te"
date: "October 6, 2014"
output: 
  ioslides_presentation:
    fig_width: 6
    fig_height: 3
---

## Comparing summary statistics

We have learned how to compute mean, median, variance, and standard deviation. We can compute these summary statistics for each of the groups (independent variables) and compare them. Let’s run through an example. There is nothing hard conceptually, fortunately.

## Iphone app study

An iPhone app company is thinking about rolling out its new user interface. They want to know whether the change will affect users’ engagement or not. They keep half of the users on the old interface and roll out the new interface to the other half. For each user, they measure engagement through time spent on the app within one session. 

## Summarize data by groups in R
Doing this in R is a bit convoluted. We use function ``ddply`` from the ``plyr`` library. Here’s an example:

```{r}
d = read.csv('http://www.typ-stats.com/datasets/ui_study.csv')
library(plyr)
ddply(d, c('age','interface'), summarize, mean=mean(duration), 
      standard_deviation=sd(duration))
```

* ``d`` is the dataset
* ``c('age', 'interface')`` are the grouping variables

## Summarize data by groups visually
This is when things get more interesting. There are many ways to do this. It depends on what you like and what point you want to make from this. We want to focus on these methods:

* A table that displays summary statistics by groups (e.g. pivot table)
* A bar graph that displays summary statistics by groups
* An overlaying histogram 
* A multi-faceted histogram 
* A boxplot 

# Summary statistics tables

## Summary statistics tables
Essentially, we want turn this R output or Pivot table into a presentable form.
R gives you something pretty raw looking.

```{r} 
library(plyr)
another_table = ddply(d, c("age","interface"), summarize, 
      mean=mean(duration), 
      standard_deviation=sd(duration),
      n=length(duration))
another_table
```

## 

You can turn it into a nicer thing.

```{r, results='asis'} 
library(knitr)
kable(another_table, format='markdown')
```

##
Or print it out and copy it to excel or something.
```{r}
write.table(another_table, sep='\t', quote=FALSE, row.names = FALSE)
```

# Summary statistics barplot

## Summary statistics barplot

Usually, we want to put the independent variables on the x-axis and the dependent variable on the y-axis.

```{r}
library(ggplot2)
ss = ddply(d, c("age","interface"), summarize, mean=mean(duration), 
           standard_deviation=sd(duration),n=length(duration))
ggplot(aes(x=interface, y=mean, fill=age),data=ss) + geom_bar(stat='identity') 
```

Yuck it is stacked up and difficult to read. The best guesses from ``ggplot`` are pretty bad. We have to tell it to be more specific.
Let’s try again

##

```{r}
ggplot(aes(x=interface, y=mean, fill=age),data=ss) + 
  geom_bar(stat='identity',position='dodge') + theme_bw(base_size=14)
```

I think this look a little better than the previous version. But it seems cramped and the color looks a bit too happy. let’s do it again.

##

```{r}
ggplot(aes(x=interface, y=mean, fill=age),data=ss) + 
  geom_bar(stat='identity', position='dodge') + 
  scale_y_continuous(name='Mean duration on app (seconds)',limits=c(0,50)) + 
  scale_fill_manual(values=c('red','blue')) + theme_bw(base_size=14)
```

It looks ok now, I think. You should keep polishing your barplot until it is beautiful to your liking.