---
title: "Unit 6"
output: 
  html_document: default
  pdf_document:
    fig_height: 3.6
    fig_width: 4.8
---

# Goals

* Boxplot
* Data visualization exercises

# Previously

We have learned how to use 

1) table
2) pivot table
3) bar plot
4) overlaying histogram
5) multi-faceted histogram. 

These are really awesome forms of data visualization. We will look at boxplot this time and look pros and cons of each method. 

# Boxplot

Histograms can be pretty messy and too fine-grained because they show so many things i.e. the count of data in each bin in each group. 

And summary statistics can be too coarse because they leave out too much information. Boxplot is something in the middle. And this is sometimes what we want.

# What is shown in a boxplot?

There is a variation for what a boxplot shows.
Traditionally, a boxplot shows six things:

1. Top whisker - some extreme high value. (we will not talk about the detail)
2. Top of the box - 75% percentile
3. Middle of the box - 50% percentile (median)
4. Bottom of the box - 25% percentile
5. Bottom whisker - some extreme low value. (we will not talk about the detail)
6. Dots - Beyond extreme values.

To carry on the UI study example from last unit, a boxplot might look like this.

For just age grouping variable, 
```{r}
d = read.csv('../datasets/ui_study.csv')
library(ggplot2)
ggplot(aes(x=age, y=duration),data=d) + geom_boxplot() + theme_bw()
```

For all interactions, 
```{r}
ggplot(aes(x=interaction(age,interface), y=duration),data=d) + geom_boxplot() + theme_bw()
```

Questions

* What would a boxplot look like if the variance is low?
* If there are many extreme values, what would a boxplot look like?
* What’s the disadvantage of boxplots compared to histogram?

\pagebreak

# Data visualization exercises

We have learned quite a few ways to compare groups. We all can use some exercises to absorb the intuition of data visualization. 

TODO: add datasets for visualizatio exercises here.